---
title: "Where the DiAna data comes from"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Where the DiAna data comes from}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

DiAna does not analyze the raw FDA Adverse Event Reporting System (FAERS)
files. It works on a cleaned version of FAERS that is rebuilt at every
quarterly update and shared on the Open Science Framework (OSF), where
`setup_DiAna()` downloads it from. This article summarizes how that data is
built and what to keep in mind when analyzing it.

The full description of the cleaning, with the code of every step, is in the
[DiAna documentation](https://github.com/fusarolimichele/DiAna_cleaning/blob/main/DiAna_Documentation.pdf)
of the [DiAna_cleaning](https://github.com/fusarolimichele/DiAna_cleaning)
repository. To cite the cleaned data: Fusaroli M, Giunchi V. DiAna:
Repository for FAERS cleaning and analysis. 2023.
[doi:10.17605/OSF.IO/ZQU89](https://doi.org/10.17605/OSF.IO/ZQU89).

## The cleaning in brief

Raw FAERS cannot be analyzed as it is: it is split into separate files for
every quarter since 2004, its format changed over time, many fields are free
text, and the same case can appear several times. The DiAna cleaning turns it
into a single, consistent database:

- **One database for all quarters.** All quarters since 2004, from both the
  legacy system (LAERS) and the current one (FAERS), are merged into one set
  of tables with the same variable names and coding throughout.
- **Standard terms.** Drug names are translated into active ingredients,
  events and indications into current MedDRA Preferred Terms, and countries,
  routes, dose forms and other free-text fields into standard values.
- **Comparable measures.** Ages are in days, weights in kilograms and
  durations in days; implausible values are set to missing rather than kept.
- **Dates kept as recorded.** Invalid dates are removed and incomplete ones are not
  completed, and the time to onset is computed for each drug.
- **One report per case.** Nullified reports, older versions of the same case
  and duplicated identifiers are removed, as are reports without any drug or
  any event.
- **Flags for selection.** Probable duplicates, premarketing (clinical trial)
  reports and literature reports are kept but flagged, so each study can
  decide whether to exclude them.
- **Ready-to-use tables.** The result is split into the tables listed below,
  which `import()` loads for each quarter.

The rest of this article describes these steps in more detail.

## The source: FAERS quarterly data

The FDA publishes FAERS as
[quarterly data extracts](https://fis.fda.gov/extensions/FPD-QDE-FAERS/FPD-QDE-FAERS.html).
DiAna uses the ASCII files, in which fields are separated by `$`. A few
things about these files shape everything that follows:

- **Quarters refer to when the FDA received a report**, not to when the
  event happened. The first quarter is January-March 2004, but reports can
  describe events that occurred earlier.
- **Each quarter is split into seven files**: DEMO (demographic and
  administrative information), DRUG (medicinal products), REAC (adverse
  events), OUTC (outcomes), RPSR (report sources), THER (therapy dates) and
  INDI (indications). From 19Q1 a DELETED file lists reports to be removed.
- **The system changed in 2012.** Until 12Q2 the data come from the Legacy
  AERS (LAERS), based on individual safety reports; from 12Q3 from FAERS,
  based on cases with numbered versions. Variable names and coding differ
  between the two, and also between FAERS quarters.
- **`primaryid` identifies one version of a case** (`caseid`). It links the
  tables: DEMO has one row per `primaryid`; DRUG, REAC, OUTC and RPSR can
  have several. THER and INDI link to DRUG through `primaryid` and
  `drug_seq`.

## How the data is cleaned

The cleaning runs in this order:

1. **Download and storage.** All quarterly ASCII files are downloaded and
   converted to R's `.rds` format. Known formatting errors are corrected
   (for example missing line breaks in three DRUG files of 2011).
2. **Merging the quarters.** For each table, all LAERS and FAERS quarters are
   merged into one. Variable names are standardized, columns never filled
   (mostly for privacy) are removed, duplicated columns are combined, empty
   strings become missing values, and each report records its quarter.
3. **Standardization** of the fields that FAERS does not code consistently:
   - **Events and indications** (`pt`, `indi_pt`, `drug_rec_act`): terms
     that are not current MedDRA Preferred Terms (older MedDRA versions, Low
     Level Terms, misspellings) are translated where possible.
   - **Drugs**: free-text drug names (brand names, active ingredients,
     abbreviations, misspellings) are translated into active ingredients
     with the
     [DiAna dictionary](https://doi.org/10.1007/s40264-023-01391-4).
     Products with several ingredients become one row per ingredient.
   - **Sex**: only `F` and `M` are kept; other values become missing.
   - **Age and weight**: converted to days and kilograms. Implausible values
     (over 122 years, over 635 kg) become missing. Age groups are recomputed
     from the age.
   - **Country**: country codes are translated into country names.
   - **Reporter occupation**: only the standard codes are kept.
   - **Dates**: dates before 1985 or after the end of the quarter become
     missing. Incomplete dates (year only, or year and month) are kept as
     they are rather than completed, to avoid inventing precision.
   - **Therapy**: durations are converted to days, and missing start or end
     dates are imputed from the duration where possible. Route, dose form,
     dose frequency, dechallenge and rechallenge are standardized.
4. **Time to onset** is computed for each drug as the days from its start
   date to the event date (day 1 being the first day of therapy).
5. **Removing reports:**
   - nullified reports listed in the DELETED files;
   - duplicated `primaryid`, keeping the most recent quarter;
   - earlier versions of the same case, keeping the most recent;
   - duplicated manufacturer report numbers, keeping the most recent;
   - reports without any drug or any event.
6. **Flagging reports**, which are kept but marked:
   - `premarketing`: reports from clinical trials, identified from the drug
     names (e.g. pre-registration codes, placebo);
   - `literature`: reports from the scientific literature;
   - `RB_duplicates` and `RB_duplicates_only_susp`: probable duplicates found
     by a rule-based algorithm (identical event date, sex, reporter country,
     age, weight, events and drugs; the second considers only suspected
     drugs). In each group of duplicates, the most recent report is not
     flagged.
7. **Splitting into the DiAna tables** listed below.

## The tables you work with

`import()` loads these tables for a quarter (the sample datasets shipped with
DiAna, such as `sample_Demo`, have the same structure):

| Table | One row per | Contents |
|---|---|---|
| `DEMO` | report | sex, age, weight, countries, dates, reporter, flags |
| `DEMO_SUPP` | report | case identifiers, version, manufacturer, quarter, report source |
| `DRUG` | report and drug | active ingredient (`substance`) and role (`role_cod`: PS, SS, I, C) |
| `DRUG_NAME` | report and drug | the drug name as reported |
| `DRUG_SUPP` | report and drug | route, dose form, dechallenge, rechallenge, lot |
| `DOSES` | report and drug | dose, unit, frequency, cumulative dose |
| `REAC` | report and event | event (MedDRA PT) and rechallenged event |
| `INDI` | report, drug and indication | indication (MedDRA PT) |
| `OUTC` | report and outcome | seriousness outcome |
| `THER` | report and drug | start and end dates, duration, time to onset |

## What to keep in mind when analyzing

- **Missing is not negative.** A report without a recorded outcome is not
  known to be non-serious; a missing sex or age is unknown. This is why
  `descriptive()` computes outcome percentages only among reports that
  record an outcome.
- **Duplicates are flagged, not removed.** To exclude probable duplicates,
  restrict the analysis, for example
  `restriction = Demo[RB_duplicates == FALSE]$primaryid` in
  `disproportionality_analysis()`.
- **Premarketing and literature reports are kept.** Exclude them with the
  `premarketing` and `literature` flags if your study requires it.
- **Not every drug name is translated.** Untranslated names still count in
  the total number of reports. Check how your drug of interest is reported
  with `get_drugnames()`, and correct the translation locally with
  `Fix_DiAna_dictionary_locally()` if needed.
- **Dates before 2013 are less reliable.** Until 2012, incomplete dates were
  completed automatically when stored, which affects time trends and time to
  onset. Null or negative times to onset are set to missing when the event
  occurred before 2013; later ones are kept, and you may want a sensitivity
  analysis excluding drugs with a null or negative time to onset.
- **Each quarter is a new release.** Analyses are reproducible only on the
  same quarter: record it (`FAERS_version`) and report it with
  `FAERS_quarter_specifics()`.

## MedDRA

The events and indications are coded with MedDRA Preferred Terms, but the
MedDRA hierarchy (HLT, HLGT, SOC) cannot be distributed with DiAna: it
requires a subscription, free for academics, from
[MedDRA](https://www.meddra.org/). To analyze at levels above the PT:

1. Download the English MedDRA release and unzip it with the password from
   the MedDRA self-service application.
2. Build `meddra_primary.csv`: one row per PT with its primary SOC, and the
   columns `def`, `soc`, `hlgt`, `hlt` and `pt`, in lowercase and separated
   by `;`. The code is in the section "Standardization > MedDRA" of the
   [DiAna documentation](https://github.com/fusarolimichele/DiAna_cleaning/blob/main/DiAna_Documentation.pdf).
3. Save it in the `external_sources` folder of your DiAna project, and load
   it with `import_MedDRA()`.

```{r meddra}
MedDRA <- import_MedDRA()
disproportionality_analysis(
  drug_selected = "paracetamol",
  reac_selected = "hepatobiliary disorders",
  meddra_level = "soc"
)
```
